Data Imputation as a Query Optimization

### 2016-12-11 paper/algorithm.tex

\input{algorithms/plan-helpers.tex}

\input{algorithms/plan-toplevel.tex}

\input{algorithms/histogram-transform.tex}

\subsection{Imputation Strategies}
The current version of the system uses an imputation strategy based on chained-equations regression trees (CERT)\cite{burgette2010multiple}. Note that plugging in a new
imputation strategy takes minimal effort, and doesn't require changes to the optimizer, so the system could in principle be tuned to a specific domain.
The imputation algorithm is iterative in nature, and proceeds by repeatedly fitting regression trees to a subset of the data and the target imputation columns.
In each iteration, the missing values of a column are replaced with newly imputed values. As the iterations proceed, the quality of the imputation
should improve as values progressively reflect more accurate relationships amongst attributes. Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

It is worth noting that given the nature of the imputation strategy used here, imputation operators become blocking in our system.

\input{algorithms/cert.tex}

\todo{Jose:Micah, take a look at the explanation above and see if it's missing anything (or is incorrect)}

>>>>

\begin{algorithm}
\input{algorithms/plan-helpers.tex}
\end{algorithm}

\begin{algorithm}
\input{algorithms/plan-toplevel.tex}
\end{algorithm}

\begin{algorithm}
\input{algorithms/histogram-transform.tex}
\end{algorithm}

\subsection{Imputation Strategies}

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data} and others) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the researcher to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2011multiple}.

These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
perform statistical analysis of the complete data. To this end, the technique of
\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
data are generated and estimators are computed by averaging over the multiple datasets. In
ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
note that in further work, multiple imputation could be used within ImputeDB to generate
lower-variance estimates of some aggregates, like averages.

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (the imputed values do not change across
epochs) or a fixed number of epochs is reached.  Algorithm~\ref{algo:imputation-strategy}
provides details on the implementation.

It is worth noting that given the nature of the imputation strategy used here, imputation
operators become blocking in our system. An extension of ImputeDB could consider regression
algorithms that learn in an online nature; this would reduce the cost of the imputation at
the possible expense of quality.

\begin{algorithm}
  \begin{algorithmic}
    \Require{$T$ is a table. $D$ is a set of attributes of $T$ that need to be imputed, and
      $C$ is a set of attributes of $T$ that has complete data }
    \Ensure{Returns an imputed $T$}

    \Function{ImputeWithCE-CART}{$H, op$}
    	\State $T' \gets \Call{ImputeRandom}{T, D}$
	\For{$n\ EPOCHS$}
		\For{$d \in D$}
            \State $imp \gets \Call{TrainRT}{\pi_{C \cup D \backslash \{d\}}(T'), \pi_d{T'}}$
            \State $T' \gets (\pi_{C \cup D \backslash \{d\}}(T'), \Call{PredictRT}{imp, T})$
		\EndFor
	\EndFor
	\Return{T'}
	\EndFunction
  \end{algorithmic}
  \caption{An algorithm for chained imputation using regression trees}
  \label{algo:imputation-strategy}
%\input{algorithms/cert.tex}
\end{algorithm}

\todo{Micah: Explanation edited a bit. In the algorithm, when we impute attribute $d$, we
condition on attributes $C \cup D \backslash \{d\}$. Hopefully the way I note it is clear;
feel free to modify otherwise.}

### 2016-12-11 paper/experiments.tex

To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic data set containing business information.\todo{Synthetic queries?}

\subsection{Data sets}

The American Community Survey provides a number of public data sets collected by the U.S. Census Bureau.
We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
The cleaning procedure is described in detail in~\cite{akande2015empirical}, but to summarize, the following data were removed:
\begin{itemize}
\item Rows corresponding to vacant houses or single occupant households.
\item Identification variables (e.g. area codes, serial numbers).
\item Flag variables.
\item Continuous variables.
\end{itemize}
The final data set consists of a single table with 671,153 rows and 35 columns, where all column entries are integers.

\todo{Describe synthetic data}

@@

We collected a set of queries that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.

The queries on the ACS data all consist only of selections, filters, and aggregations.
The ACS data is contained in a single table, so there are relatively few join queries that are interesting on this data set.
However, even with select, filter, and aggregate, there are interesting choices to make with imputation placement.

\begin{figure}
  \centerfloat
  \input{queries}
  \caption{The queries used in our experiments.}
\end{figure}

\begin{figure}
  \centerfloat
  \input{runtimes}
  \caption{Running time and error for queries with different imputation levels.}
\end{figure}

>>>>

To evaluate the performance of our system, we generated plans for queries on two data sets: the American Community Survey and a synthetic dataset.
\subsection{Data sets}

The American Community Survey (ACS) provides a number of public data sets collected by the U.S. Census Bureau.
We used a cleaned version of the 2012 Public Use Microdata Sample (PUMS) data kindly provided by the authors of~\cite{akande2015empirical}.
The cleaning procedure is described in detail in~\cite{akande2015empirical}, but to summarize, the following data were removed:
\begin{itemize}
\item Rows corresponding to vacant houses or single occupant households.
\item Identification variables (e.g. area codes, serial numbers).
\item Flag variables.
\item Continuous variables.
\end{itemize}
The final data set consists of a single table with 671,153 rows and 37 columns, where all column entries are integers.

In order to evaluate our implementation, we created a dirtied variant of the ACS data by randomly eliminating
10\% of the fields across the table.

In order to evaluate additional queries, we constructed a synthetic table, consisting of 10 columns and 10,000 rows, with
all values drawn from a uniform distribution between 0 and 100. To create a dirtied variant of the synthetic data, we randomly deleted 30\% of the fields
across the table.

@@

We collected a set of queries (Figure~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.

The queries on the ACS data all consist only of selections, filters, and aggregations.
The ACS data is contained in a single table, so there are relatively few join queries that are interesting on this data set and remain relevant for analysts, rather
than contrived solely for the purpose of experimentation.
However, even with select, filter, and aggregate, there are interesting choices to make with imputation placement. In order to explore joins, we leverage the synthetic
dataset and write various ad-hoc queries.

\begin{figure}
  \centerfloat
  \input{queries}
  \caption{The queries used in our experiments.}
  \label{fig:queries}
\end{figure}

\begin{figure}
  \centerfloat
  \input{runtimes}
  \caption{Running time and error for queries with different imputation levels.}
  \label{fig:experiments}
\end{figure}

Figure~\ref{fig:experiments} provides a summary of our experimental results. For purposes of simplicity, we present only the results associated with
queries that return a result size matching our base case. This addresses the fact that deriving a reasonable error metric for set-queries 
is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of errors squared
(SSE). Figure~\ref{fig:experiments} indicates the query number and the error produced by a query engine with traditional SQL null semantics (base error).
For imputations, we present two extremes of the $\alpha$ parameter: 0 and 1, which reflect a a focus on runtime and information loss, respectively.
For each query, the table displays the new error ($\epsilon$), the planning time ($t_p$) and the running time ($t_r$) in seconds.

The overall low planning times highlight the practicality of our algorithm, despite the exponential complexity. Runtimes clearly vary based on $\alpha$, allowing
users to tune their query performance as desired.

\todo{add result-specific analysis}

Finally, we performed CERT imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
We note that the latter took \todo{add time}, compared to our maximum of \todo{add time}. Given the varying needs of database users, \ProjectName{} clearly allows flexibility without
a significant performance hit.
